---
title: "About this blog"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Welcome to our blog! Here you'll find the latest news, insights, and examples of doing deep learning (using [torch](https://github.com/mlverse/torch), [TensorFlow](https://github.com/rstudio/tensorflow) and [Keras](https://github.com/rstudio/keras)), distributed computing and ML automation (using [sparklyr](https://github.com/sparklyr/sparklyr) and [mlflow](https://github.com/mlflow/mlflow)), and data ingestion (with [pins](https://github.com/rstudio/pins)) from R.

We also like to draw from a wide variety of contributors (so please consider submitting a post if so motivated!). Here is an alphabetic list of authors so far:

| Author             | Affiliation        | GitHub                             |
|--------------------|--------------------|------------------------------------|
| Turgut Abdullayev  | QSS Analytics      | <https://github.com/henry090>      |
| J.J. Allaire       | RStudio            | <https://github.com/jjallaire>     |
| François Chollet   | Google Brain       | <https://github.com/fchollet>      |
| Matt Dancho        | Business Science   | <https://github.com/mdancho84>     |
| Daniel Falbel      | RStudio            | <https://github.com/dfalbel>       |
| Leon Eyrich Jessen | Tech. U of Denmark | <https://github.com/leonjessen>    |
| Sigrid Keydana     | RStudio            | <https://github.com/skeydan>       |
| Yitao Li           | RStudio            | <https://github.com/yitao-li>      |
| Javier Luraschi    | RStudio            | <https://github.com/javierluraschi>|
| Pete Mohanty       | Stanford           | <https://github.com/rdrr1990>      |
| Thomas Lin Pedersen| RStudio            | <https://github.com/thomasp85>     |
| Nick Strayer       | Vanderbilt         | <https://github.com/nstrayer/>     |
| Yuan Tang          | Ant Group          | <https://github.com/terrytangyuan> |


If you are interested in publishing some of your own work or insights on the RStudio AI Blog, we'd love to hear from you! See the article on [contributing posts](contributing.html) for additional details.

The RStudio AI Blog was created using [Distill for R Markdown](https://rstudio.github.io/distill).

<br/>

```{r, results="asis"}
cat(paste(readLines("_subscribe.html", warn = FALSE), collapse = "\n"))
```
