---
title: "About this blog"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


Welcome to our blog! Here you'll find the latest news, insights, and examples of using [TensorFlow for R](https://tensorflow.rstudio.com). In particular, we like to publish lots of in-depth case studies that go beyond the more straightforward TensorFlow examples you often find online.

We also like to draw from a wide variety of contributors (so you should consider submitting a post if so motivated!). Here are some of the authors that have written articles so far:

| Author  | Affiliation | GitHub  |
|---|---|---|
| J.J. Allaire  | RStudio | https://github.com/jjallaire  |
| François Chollet  | Google Brain | https://github.com/fchollet  |
| Sigrid Keydana  | RStudio | https://github.com/skeydan  |
| Daniel Falbel  | Curso-R | https://github.com/dfalbel  |
| Matt Dancho  | Business Science | https://github.com/mdancho84  |
| Yuan Tang | Ant Financial | https://github.com/terrytangyuan |
| Pete Mohanty | Stanford | https://github.com/rdrr1990 |
| Leon Eyrich Jessen | Tech. U of Denmark | https://github.com/leonjessen |
| Nick Strayer | Vanderbilt | https://github.com/nstrayer/ |

If you are interested in publishing some of your own work or insights on the TensorFlow for R blog, we'd love to hear from you! See the article on [contributing posts](contributing.html) for additional details.

The TensorFlow for R blog was created using [Radix for R Markdown](https://rstudio.github.io/radix).

<br/>

```{r, results="asis"}
cat(paste(readLines("_subscribe.html"), collapse = "\n"))
```

