---
title: "TensorFlow for R Blog"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


Welcome to our blog! Here you'll find the latest news, insights, and examples of using [TensorFlow for R](https://tensorflow.rstudio.com). 

In particular, we like to publish lots of in-depth case studies that go beyond the more straightforward TensorFlow examples you often find online. Whenever possible we cover data preparation, visualization, and comparison to conventional methods to make sure that neural networks are even a good fit for the problem domain! 

We also like to draw from a wide variety of contributors (so you should consider submitting a post if so motivated!). Here are some of the authors that have written articles so far:

| Author  | Affiliation | GitHub  |
|---|---|---|
| J.J. Allaire  | RStudio | https://github.com/jjallaire  |
| François Chollet  | Google Brain | https://github.com/fchollet  |
| Sigrid Keydana  | RStudio | https://github.com/skeydan  |
| Daniel Falbel  | Curso-R | https://github.com/dfalbel  |
| Matt Dancho  | Business Science | https://github.com/mdancho84  |
| Yuan Tang | Ant Financial | https://github.com/terrytangyuan |
| Pete Mohanty | Stanford | https://github.com/rdrr1990 |
| Leon Eyrich Jessen | Tech. U of Denmark | https://github.com/leonjessen |

If you are interested in publishing some of your own work or insights on the TensorFlow for R blog, we'd love to hear from you! See the article on [contributing posts](contributing.html) for additional details.

<br/>

```{r, results="asis"}
cat(paste(readLines("_subscribe.html"), collapse = "\n"))
```

<!---
TODO: Contributing page
TODO: Add disqus
TODO: Add google analytics

TODO: Redirect posts/feed from original blog
--->
