---
title: "Training ImageNet with R"
description: >
  This post explores how to train large-scale datasets with TensorFlow and R. Specifically, we present how to download and repartition ImageNet, followed by training these partitions across multiple GPUs in distributed environments
author:
  - name: Javier Luraschi
    affiliation: RStudio
    affiliation_url: https://www.rstudio.com/
date: 08-24-2020
categories:
  - R
  - TensorFlow/Keras
  - Distributed Computing
  - Data Management
bibliography: bibliography.bib
preview: images/fishing-net.jpg
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

[ImageNet](http://www.image-net.org/) [@deng2009imagenet] is an image database organized according to the [WordNet](http://wordnet.princeton.edu/) [@miller1995wordnet] hierarchy which, historically, has been used in Computer Vision benchmarks and research. However, it was not until AlexNet [@krizhevsky2012imagenet] introduced Deep Learning concepts like Convolutional Neural Networks and GPUs, that the Computer Vision discipline turned to Deep Learning to achieve state-of-the-art models that revolutionized their field. Given the importance of ImageNet and AlexNet, we want to use this post to introduce the tools annd techniques that we can consider when training ImageNet and other large-scale datasets from R.

Now, in order to process the ImageNet dataset, we will first have to -- divide and conquer -- by paratitioning into several manegeable subsets. Afterwards, we will train ImageNet using AlexNet across multiple GPUs and compute instances. [Preprocessing ImageNet](#preprocessing-imagenet) and [distributed training](#distributed-training) are the two topics that this post will present and discuss, strating with preprocessing ImageNet.

## Preprocessing ImageNet

When dealing with large datasets, even simple tasks like downloading or reading a dataset can be much harder than what you would expect. For instance, since ImageNet is roughly 300GB in size, you will need to make sure we have at least 600GB of free space to leave some room for download and decompression. I personally don't have 600GB to spare, not even a 600GB disk drive at all! But no worries, you can always borrow a reasonable machine with a huge disk drive from your favorite cloud provider. While you are at it, you should also request compute instances with multiple GPUs, Solid State Drives (SSDs), and a reasonable amount of CPUs and memory. If you want to use exact configuration we used, take a look at the [mlverse/imagenet](https://github.com/mlverse/imagenet) repo, which contains a Docker image and configuration commands required to provision reasonable computing resources for this task. Now, we don't want to put too much emphasis into this resource-allocation step since you are likely to use a different cloud provider, or be lucky enough to have access to similar resources already. In summary, make sure you have access to sufficient compute resources.

Now that we have resources capable of working with ImageNet, we need to find a place to download ImageNet from. Long story short, the easiest way is to use a variation of ImageNet used in the ImageNet Large Scale Visual Recognition Challenge (ILSVRC), which contains a subset of about 250GB of data and can be easily downloaded from their [Kaggle](https://kaggle.com) competition.

If you've read some of our previous posts, you might be thinking of using -- the [pins](https://pins.rstudio.com) package! The pins package enables you to cache, discover and share resources from many services, including Kaggle. You can learn more about data retrieval from Kaggle in the [Using Kaggle Boards](http://pins.rstudio.com/articles/boards-kaggle.html) article; in the meantime, lets assume we are familiar with this package.

All we need to do now is register the Kaggle board, retrieve ImageNet as a pin, and uncompress this file. Warning, the following code requires you to stare at a progress bar for, potentially, over an hour!

```{r}
library(pins)
board_register("kaggle", token = "kaggle.json")

pin_get("c/imagenet-object-localization-challenge", board = "kaggle")[1] %>%
  untar(exdir = "/localssd/imagenet/")
```

If we are going to be training this model over-and-over using multiple GPUs and even multiple compute instances, we want to make sure we don't waste too much time downloading ImageNet every single time.

The first improvement to consider is getting a faster hard drive. In our case, we locally-mounted an array of SSDs into the `/localssd` path. We then used `/localssd` to extract ImageNet and configured R's temp paths and pins cache to use the SSDs as well. Consult your cloud provider documentation to configure SSDs or take a look at [mlverse/imagenet](https://github.com/mlverse/imagenet).

Next, a well-known approach we can follow is to, partition ImageNet into chunks that can be individually downloaded to perform distributed training later on. 

In addition, is also faster to download ImageNet from a near-by location, ideally from a URL stored within the same datacenter where our cloud instance is located. For this, we can also use pins to register a board with our cloud-provider and then re-upload each partition. Since ImageNet is already partitioned by category, we can easily split ImageNet into multiple zip files and re-upload to your closest datacenter as follows. Make sure the storage bucket is created in the same region as your compute instances!

```{r}
board_register("<board>", name = "imagenet", bucket = "r-imagenet")

train_path <- "/localssd/imagenet/ILSVRC/Data/CLS-LOC/train/"
for (path in dir(train_path, full.names = TRUE)) {
  dir(path, full.names = TRUE) %>%
    pin(name = basename(path), board = "imagenet", zip = TRUE)
}
```

We can now retrieve a subset of ImageNet quite efficiently. If you are motivated to do so and have about one gigabyte to spare, feel free to follow along executing this code. Notice that ImageNet contains -- lots of JPEG images for each WorNet category.

```{r}
board_register_datatxt("https://storage.googleapis.com/r-imagenet/", "imagenet")

categories <- pin_get("categories", board = "imagenet")
pin_get(categories$id[1], board = "imagenet", extract = TRUE) %>%
  tibble::as_tibble()
```
```
# A tibble: 1,300 x 1
   value                                                           
   <chr>                                                           
 1 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10026.JPEG
 2 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10027.JPEG
 3 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10029.JPEG
 4 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10040.JPEG
 5 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10042.JPEG
 6 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10043.JPEG
 7 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10048.JPEG
 8 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10066.JPEG
 9 /home/rstudio/.cache/pins/storage/n01440764/n01440764_10074.JPEG
10 /home/rstudio/.cache/pins/storage/n01440764/n01440764_1009.JPEG 
# … with 1,290 more rows
```

When doing distributed training over ImageNet, we can now let a single compute instance process a partition of ImageNet with ease. Say, 1/16 of ImageNet can be retrieved and extracted, in under a minute, using parallel downloads with the [callr](https://callr.r-lib.org/) package:

```{r}
categories <- pin_get("categories", board = "imagenet")
categories <- categories$id[1:(length(categories$id) / 16)]

procs <- lapply(categories, function(cat)
  callr::r_bg(function(cat) {
    library(pins)
    board_register_datatxt("https://storage.googleapis.com/r-imagenet/", "imagenet")
    
    pin_get(cat, board = "imagenet", extract = TRUE)
  }, args = list(cat))
)
  
while (any(sapply(procs, function(p) p$is_alive()))) Sys.sleep(1)
```

Last but not least, we can create a `tibble` to hold categories and image paths in a format that is compatible with [tfdatasets](https://tensorflow.rstudio.com/guide/tfdatasets/introduction/).

```{r}
data <- list(
    image = unlist(lapply(categories, function(cat) {
        pin_get(cat, board = "imagenet", download = FALSE)
    })),
    category = unlist(lapply(categories, function(cat) {
        rep(cat, length(pin_get(cat, board = "imagenet", download = FALSE)))
    })),
    categories = categories
)
```

Great! We are halfway there training ImageNet. The next section will focus on introducing distributed training using multiple GPUs.

## Distributed Training

Now that we have broken down ImageNet into manegeable parts, we can forget for a second about the size of ImageNet and focus on training a Deep Learning model for this dataset. However, any model we choose is likely to require a GPU, even for a 1/16 subset of ImageNet. So make sure your GPUs are properly configured by running `is_gpu_available()`. If you need help getting a GPU configured, the [Using GPUs with TensorFlow and Docker](https://www.youtube.com/watch?v=i5Bjm3jG_d8) video can help you get up to speed.

```{r}
library(tensorflow)
tf$test$is_gpu_available()
```
```
[1] TRUE
```

We can now choose which Deep Learning model would be best suited for ImageNet classification tasks. Instead, for this post, we will go back in time to the glory days of AlexNet and use the [r-tensorflow/alexnet](https://github.com/r-tensorflow/alexnet) repo instead. This repo contains a port of AlexNet to R, but please notice that this port has not been tested and is not ready for any real use cases. In fact, we would appreciate PRs to improve it if someone feels inclined to do so. Regardless, the focus of this post is on workflows and tools, not about achieving state-of-the-art image classification scores. So by all means, feel free to use more appropriate models.

Once we've choosen a model, we will want to me make sure that it properly trains on a subset of ImageNet:

```{r}
remotes::install_github("r-tensorflow/alexnet")
alexnet::alexnet_train(data = data)
```
```
Epoch 1/2
 103/2269 [>...............] - ETA: 5:52 - loss: 72306.4531 - accuracy: 0.9748
```

So far so good! However, this post is about enabling large-scale training across multiple GPUs, so we want to make sure we are using as many as we can. Unfortunately, running `nvidia-smi` will show that only one GPU currently being used:

```{bash eval=FALSE}
nvidia-smi
```
```
+-----------------------------------------------------------------------------+
| NVIDIA-SMI 418.152.00   Driver Version: 418.152.00   CUDA Version: 10.1     |
|-------------------------------+----------------------+----------------------+
| GPU  Name        Persistence-M| Bus-Id        Disp.A | Volatile Uncorr. ECC |
| Fan  Temp  Perf  Pwr:Usage/Cap|         Memory-Usage | GPU-Util  Compute M. |
|===============================+======================+======================|
|   0  Tesla K80           Off  | 00000000:00:05.0 Off |                    0 |
| N/A   48C    P0    89W / 149W |  10935MiB / 11441MiB |     28%      Default |
+-------------------------------+----------------------+----------------------+
|   1  Tesla K80           Off  | 00000000:00:06.0 Off |                    0 |
| N/A   74C    P0    74W / 149W |     71MiB / 11441MiB |      0%      Default |
+-------------------------------+----------------------+----------------------+
                                                                               
+-----------------------------------------------------------------------------+
| Processes:                                                       GPU Memory |
|  GPU       PID   Type   Process name                             Usage      |
|=============================================================================|
+-----------------------------------------------------------------------------+
```

In order to train across multiple GPUs, we need to define a distributed-processing strategy. If this is a new concept, it might be a good time to take a look at the [Distributed Training with Keras](https://tensorflow.rstudio.com/tutorials/advanced/distributed/distributed_training_with_keras/) tutorial and the [distributed training with TensorFlow](https://www.tensorflow.org/guide/distributed_training) docs. Or, if you allow us to overly-simplify the process, all you have to do is define and compile your model under the right scope. A step-by-step explanation is available in the [Distributed Deep Learning with TensorFlow and R](https://www.youtube.com/watch?v=DQyLTlD1IBc) video. In this case, the `alexnet` model [already supports](https://github.com/r-tensorflow/alexnet/blob/57546/R/alexnet_train.R#L92-L94) a strategy parameter, so all we have to do is pass it along.

```{r  eval=FALSE}
library(tensorflow)
strategy <- tf$distribute$MirroredStrategy(
  cross_device_ops = tf$distribute$ReductionToOneDevice())

alexnet::alexnet_train(data = data, strategy = strategy, parallel = 6)
```

Notice also `parallel = 6` which configures `tfdatasets` to make use of multiple CPUs when loading data into our GPUs, see [Parallel Mapping](https://tensorflow.rstudio.com/guide/tfdatasets/introduction/#parallel-mapping) for details.

We can now re-run `nvidia-smi` to validate all our GPUs are being used!

```{bash eval=FALSE}
nvidia-smi
```
```
+-----------------------------------------------------------------------------+
| NVIDIA-SMI 418.152.00   Driver Version: 418.152.00   CUDA Version: 10.1     |
|-------------------------------+----------------------+----------------------+
| GPU  Name        Persistence-M| Bus-Id        Disp.A | Volatile Uncorr. ECC |
| Fan  Temp  Perf  Pwr:Usage/Cap|         Memory-Usage | GPU-Util  Compute M. |
|===============================+======================+======================|
|   0  Tesla K80           Off  | 00000000:00:05.0 Off |                    0 |
| N/A   49C    P0    94W / 149W |  10936MiB / 11441MiB |     53%      Default |
+-------------------------------+----------------------+----------------------+
|   1  Tesla K80           Off  | 00000000:00:06.0 Off |                    0 |
| N/A   76C    P0   114W / 149W |  10936MiB / 11441MiB |     26%      Default |
+-------------------------------+----------------------+----------------------+
                                                                               
+-----------------------------------------------------------------------------+
| Processes:                                                       GPU Memory |
|  GPU       PID   Type   Process name                             Usage      |
|=============================================================================|
+-----------------------------------------------------------------------------+
```

The `MirroredStrategy` can help us scale up to about 8 GPUs per compute instance; however, we are likely to need 16 instances with 8 GPUs each to train ImageNet in a resonable time, see Jeremy Howard's post on [Training Imagenet in 18 Minutes](https://www.fast.ai/2018/08/10/fastai-diu-imagenet/). So where do we go from here?

Welcome to `MultiWorkerMirroredStrategy`, this strategy can use not only multiple GPUs, but also multiple GPUs across multiple computers. To configure them, all we have to do is specify to TensorFlow a `TF_CONFIG` of addresses and run the exact same code in each computer.

```{r eval=FALSE}
library(tensorflow)

Sys.setenv(TF_CONFIG = jsonlite::toJSON(list(
    cluster = list(
        worker = c("10.100.10.100:10090", "10.100.10.101:10091")
    ),
    task = list(type = 'worker', index = 0)
), auto_unbox = TRUE))

strategy <- tf$distribute$MultiWorkerMirroredStrategy(
  cross_device_ops = tf$distribute$ReductionToOneDevice())

alexnet::alexnet_train(data = data, strategy = strategy, parallel = 6)
```

Please note that `index = 0` must change for each compute instance to uniquely identify it. In addition, `data` should point to a different partition of ImageNet, which we can easily retrieve with `pins`. Other than that, the code is exactly the same when running this distributed.

However, it can be quite time-consuming and error-prone to manually run code in each R session. So instead, we should think of making use of tools like Apache Spark and [Barrier Execution](https://blog.rstudio.com/2020/01/29/sparklyr-1-1/#barrier-execution) to easily run the same code across multiple nodes -- without having to manually run code across multiple R instances. This would require that we also set up Apache Spark, which goes beyond the scope of this post, but feel free to watch the [Deep Learning with Spark, TensorFlow and R](https://www.youtube.com/watch?v=Zm20P3ADa14) video to learn more about this.

We hope this post gave you a complete overview of what training large-datasets in R looks like -- Thanks for reading along!
